Semantic Linearity in a Lazy Optimising Compiler

### 2024-01-24 icfp.tex
notes

\title{Type-checking Linearity in GHC Core}
\subtitle{Semantic Linearity in a Lazy Optimising Compiler}

\author{Rodrigo Mesquita}
\affiliation{
  \department{Departamento de Inform\'{a}tica}
  \institution{FCT-NOVA, Universidade Nova de Lisboa}
  %\city{City}
  \country{Portugal}}
\affiliation{
   \institution{Well-Typed}
  %\city{City}
   \country{Something}}
 \email{xxx@yyy.com}
\author{Bernardo Toninho}
\orcid{0000-0002-0746-7514}
\affiliation{
  \department{Departamento de Inform\'{a}tica}
  \institution{NOVA-LINCS, FCT-NOVA, Universidade Nova de Lisboa}
  %\city{City}
  \country{Portugal}}
\email{btoninho@fct.unl.pt}
%% TODO: AFFILIATION, EMAIL
%% TODO: CCS CONCEPTS and KEYWORDS
%% TODO: ACM RIGHTS notice...
% CHECK: https://www.acm.org/binaries/content/assets/publications/taps/latex-best_practices-06-may-2020.pdf

@@

To explore these case properties in the presence of linearity, we start with an
example of a program that constructs a tuple from linear resources then pattern
matches on it, then uses both linearly-bound variables from the tuple pattern
match. This is well-typed in Linear Haskell:
\begin{working}
\begin{lstlisting}
f11 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* b *' $\lolli$ '* c) -> c
f11 x y use = case (x,y) of { (a,b) -> use a b }
\end{lstlisting}
\end{working}
What might be more surprising is that a similar program which discards the
pattern variables and instead uses the resources in the scrutinee is also
semantically linear, despite not being accepted by Linear Haskell:
\begin{notyet}
\begin{lstlisting}
f12 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* b *' $\lolli$ '* c) -> c
f12 x y use = case (x,y) of z { (a,b) -> use x y }
\end{lstlisting}
\end{notyet}
We justify that \incode{f12} is linear by appealing to the fact that the
expression (the tuple) being scrutinized is already in WHNF, evaluating it will
not consume neither $x$ nor $y$. Even if the tuple was constructed with
two expressions using $x$ and $y$ respectively, no computation would happen
since we aren't using neither $a$ nor $b$ (thereby never forcing the arguments
of the tuple). However, if we did use $a$ in the case body, then $x$ would be unavailable:
\begin{noway}
\begin{lstlisting}
f13 :: a *' $\lolli$ '* a *' $\lolli$ '* (a *' $\lolli$ '* a *' $\lolli$ '* c) -> c
f13 x y use = case (x,y) of z { (a,b) -> use a x }
\end{lstlisting}
\end{noway}
This idea that $x$ and $a$ are mutually exclusive is the same behind let
bindings being mutually exclusive to the resources that define them.
By forcing the pattern variable (or the let binding), we run the computations
defined in terms of the linear variables used for that constructor argument (or
let binder body), but otherwise, if we don't use those binders, then we don't
run the computation thus no resources are consumed.

@@

Dually, consider the scrutinee to be an expression that's not in WHNF, s.t.
evaluating it to WHNF will require doing computation and thus consume linear
resources that are used in it:
\begin{notyet}
\begin{lstlisting}
f15 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* b *' $\lolli$ '* (c,d)) -> (c,d)
f15 x y use = case use x y of z { (a,b) -> z }
\end{lstlisting}
\end{notyet}
Unlike when the scrutinee was in WHNF, we can no longer use $x,y$ in the case
alternatives, but we \emph{must} still use either the case binder $z$ or the linear
pattern variables $a,b$, e.g. it would be quite disastrous if any of the following typechecked:
\begin{noway}
\begin{lstlisting}
doubleFree :: Ptr *' $\lolli$ '* (Ptr *' $\lolli$ '* Result) -> Result
doubleFree x free = case free x of z { Result v -> free x }
\end{lstlisting}
\end{noway}
\begin{noway}
\begin{lstlisting}
leakPointer :: Ptr -> ()
leakPointer x = case id x of z { _ -> () }
\end{lstlisting}
\end{noway}
%
%whereas this is fine:
%\begin{notyet}
%\begin{code}
%f16 :: Ptr -o (Ptr -o (Value,Int)) -> (Value,Int)
%f16 x free = case K (free x) of z { K y -> free x }
%\end{code}
%\end{notyet}
%
The result of evaluating the scrutinee must be consumed exactly to guarantee
that the resources used in the scrutinee are fully consumed, or risk them being only ``almost'' consumed. Take for example
\incode{use} in \incode{f15} to simply be \incode{(,)}: it is not sufficient for \incode{use x y} to be
evaluated to WHNF to consume \incode{x} and \incode{y}. Otherwise, if all the resources were
considered to be fully consumed after the scrutinee were evaluated in a case
expression, we could simply ignore the pattern variables, effectively
discarding linear resources (for cases such as the \incode{use = (,)} example). In
short, if the scrutinee is not in WHNF we must either consume the case binder
or the linear components of the pattern.

@@

\begin{figure}[t]
  \begin{minipage}{0.5\textwidth}
\begin{noway}
\begin{lstlisting}
f w = case w of z
        (a,b) -> case (a,b) of z'
                        (c,d) -> (a,c)
\end{lstlisting}
\end{noway}
\vspace{-0.5cm}
\caption{Bad Case\label{fig:badcase}}
\end{minipage}%
\begin{minipage}{0.5\textwidth}
\begin{notyet}
\begin{lstlisting}
f w = case w of z
        (a,b) -> case (a,b) of z'
                        (c,d) -> (a,d)
\end{lstlisting}
\end{notyet}
\vspace{-0.5cm}
\caption{Good Case\label{fig:goodcase}}
\end{minipage}
\end{figure}

@@

\section{Linear Core as a GHC Plugin\label{sec:discuss:implementation}}

@@

The usage environments of a recursive group of binders (that is not necessarily
strongly-connected) is computed in two separate passes.
%
First, we calculate a \emph{naive environment} by recording the linear
resources semantically used in the binder bodies, while counting \emph{uses}
(not syntactic occurrences) of the mutually recursive let variables.
%
Second, the binder names and corresponding \emph{naive environment} (mapping
each linear resource to $1$ and the recursive variables to the $n$ number of
times they are \emph{used}) in Cartesian pairs are given as input to
% Algorithm~\ref{computeRecUsages},
an algorithm which computes the actual usage environment
of each binder.
%
% TODO: I should probably use the re-computed usageEnvs instead of the naiveUsageEnvs.
% I'm pretty sure it might fail in some inputs if I keep using the naiveUsageEnvs.
% \begin{algorithm}
% $usageEnvs \gets naiveUsageEnvs.map(fst)$\;
% \For{$(bind, U) \in naiveUsageEnvs$}{
%     \For{$V \in usageEnvs$}{
%         $V \gets sup(V[bind]*U\setminus\{bind\}, V\setminus\{bind\})$
%     }
% }
% \caption{computeRecUsages\label{computeRecUsages}}
% \end{algorithm}
%
Intuitively, the algorithm, for each recursive binder, iterates over the
(initially naive) usage environments and substitutes occurrences of the
recursive binders by their corresponding usage environment,
scaled up by the amount of times that recursive binder is used in the
environment being updated. The result is the least upper bound of linear
resources used by each strongly-connected group of mutually recursive bindings
(for each binder of such a group), since all occurrences of the recursive
binders in the usage environments will be substituted away by the corresponding
recursive usage environments.
%
% The high-level description is:
% \begin{itemize}
%     \item Given a list of binders and their naive environment ($(f, F), (g, G),
%     (h, H)$) in which each use of $f, g, h$ is mapped to a count $n$ in $F, G, H$
%
%     \item For each pair of a letrec-bound variable and corresponding usage
%     environment, update all bindings and their usage as described in the
%     algorithm
%
%     \item After iterating through all bound rec vars, all usage environments
%         will be free of recursive bind usages, and hence describe the ``final'' usage environment
% \end{itemize}
%
The complexity of computing a usage environment for $n$ recursive let binders
is quadratic in $n$,
% i.e. the algorithm has $O(n^2)$ complexity,
but this is not an issue since it is uncommon to have more than a handful of
binders in the same recursive let block.

@@

Programs ``linear modulo call-by-name'' are a class of programs
which scrutinize a variable, but then uses the variable in the case
alternatives of the case scrutinizing it. As will be made clear in
Section~\ref{sec:reverse-binder-swap-considered-harmful}, these programs can be
understood as linear as long as applications of linear functions binding these
variables are \emph{not} reduced \emph{call-by-name}, because, in doing so,
linear resources are duplicated. In Core, these programs are not seen as
linear, and thus GHC does not not reduce them using a call-by-name evaluation strategy.
% We have a flag in our implementation to accept some of these programs...
%
``Linear Rejected'' programs include different kinds of programs which are
rejected by the Linear Core system, but are still semantically linear. An
example is a program to which a rewrite rule was applied, resulting in an
application of unrestricted function to a linear resource in the first argument
to \incode{build}:
\begin{lstlisting}
take :: Int -> Replicator a *' $\lolli$ '* [a]
take = \ (ds :: Int) (r :: Replicator b) ->
     case ds of
       1 -> build (\(c :: b -> a -> a) (n :: a) -> c (extract r) n)
       ...
\end{lstlisting}
We know the linear resource is still used linearly because \incode{build} will
always instance the unrestricted function to $(:)$ (read \emph{Cons}), which is
linear. Additionally, these include programs scrutinizing a value of a type of
which all constructors have no linear components, but only matching on the
default alternative, programs where common-sub-expression elimination
substituted more than one occurrence of \incode{Ur x} by \incode{y}, in the body of
scrutinizing \incode{y} (a linear variable), and other programs affected by rewrite rules.
%
``Not Linear Rejected'' indicate programs that we do not understand as linear,
semantically, and are simultaneously rejected by Linear Core and its
implementation. An example, where the last component of \incode{HashMap}, \incode{wwz},
is linear, but is not being consumed in the case alternative matching on it:
\begin{lstlisting}
jssvi :: Ur Bool *' $\lolli$ '* Set Int *' $\lolli$ '* Ur (TestT IO ())
jssvi (a :: Ur Bool) (b :: Set Int)
  = case a of
     Ur ss -> case b of { HashMap wwx wwy wwz -> jump wjssAd ss }
\end{lstlisting}
%
Finally, ``Unknown Rejected'' programs are those whose validity we did not
check. These include both programs accepted by Linear Core, but not by its
implementation, and programs that are simply rejected by Linear Core, but which
were not categorized.

@@

\section{Conclusion\label{sec:discussion}}

Linear Core is an intermediate language with a type system system that
understands (semantic) linearity in the presence of laziness, suitable for
optimising compilers with these characteristics which leverage laziness and
(possibly) linearity in its transformations.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% {{{ Related Work
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Related Work}

In this section we discuss related work, namely, Linear
Haskell~\cite{cite:linearhaskell}, Linear Mini-Core~\cite{cite:minicore}, and
linearity-influenced optimising
transformations~\cite{cite:let-floating,peytonjones1997a,cite:linearhaskell}.

@@

\paragraph{Linear Haskell\label{sec:related-work-linear-haskell}}

% \todo[inline]{Say something about Linear Haskell's lazy operational semantics,
% but the type system not being concerned with linearity in the presence of
% laziness}

Haskell, contrary to most programming languages with linear types, has existed
for 31 years of its life \emph{without} linear types. As such, the introduction
of linear types to Haskell comes with added challenges that do not exist in
languages that were designed with linear types from the start:
%
\begin{itemize}
    \item Backwards compatibility. The addition of linear types shouldn't break
        all existing Haskell code.
    \item Code re-usability. The linearly-typed part of Haskell's ecosystem and
        its non-linearly-typed counterpart should fit in together and it must be
        possible to define functions readily usable by both sides
        simultaneously.
    \item Future-proofing. Haskell, despite being an
        industrial-strength language, is also a petri-dish for experimentation
        and innovation in the field of programming languages. Therefore, Linear
        Haskell takes care to accommodate possible future features, in
        particular, its design is forwards compatible with affine and dependent
        types.
\end{itemize}
%
Linear Haskell~\cite{cite:linearhaskell} is thus concerned with retrofitting
linear types in Haskell, taking into consideration the above design goals, but
is not concerned with extending Haskell's intermediate language(s),
which presents its own challenges.

Nonetheless, while the Linear Haskell work keeps Core unchanged, its
implementation in GHC does modify and extend Core with linearity/multiplicity
annotations. Core's type system is unable to type \emph{semantic} linearity of
programs (in contrast to \emph{syntactic} linearity), which results in Core
rejecting linear programs resulting from optimising transformations that
leverage the non-strict semantics of Core.
%
Linear Core overcomes the limitations of Core's linear type system derived from
Linear Haskell by understanding semantic linearity in the presence of laziness,
and provably accepts multiple Core-to-Core passes. Linear Core, ultimately, can
also be seen as a system that validates the programs written in Linear Haskell
and are compiled by GHC, by guaranteeing (through typing) that linear resources
are still used exactly once throughout the optimising transformations.

>>>>

\title{Semantic Linearity in a Lazy Optimising Compiler}
\subtitle{Type-checking Linearity in GHC Core}

\author{Rodrigo Mesquita}
\affiliation{
  \department{Departamento de Inform\'{a}tica}
  \institution{FCT-NOVA, Universidade Nova de Lisboa}
  %\city{City}
  \country{Portugal}}
\affiliation{
   \institution{Well-Typed LLP}
  %\city{City}
  %\country{Something}
 }
 \email{xxx@yyy.com}
\author{Bernardo Toninho}
\orcid{0000-0002-0746-7514}
\affiliation{
  \department{Departamento de Inform\'{a}tica}
  \institution{NOVA-LINCS, FCT-NOVA, Universidade Nova de Lisboa}
  %\city{City}
  \country{Portugal}}
\email{btoninho@fct.unl.pt}
%% TODO: AFFILIATION, EMAIL
%% TODO: CCS CONCEPTS and KEYWORDS
%% TODO: ACM RIGHTS notice...
% CHECK: https://www.acm.org/binaries/content/assets/publications/taps/latex-best_practices-06-may-2020.pdf

@@

To explore these case properties in the presence of linearity, we start with an
example of a program that constructs a tuple from linear resources then pattern
matches on it, then uses both linearly-bound variables from the tuple pattern
match. This is well-typed in Linear Haskell:
\begin{working}
\begin{lstlisting}
f11 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* b *' $\lolli$ '* c) -> c
f11 x y use = case (x,y) of { (a,b) -> use a b }
\end{lstlisting}
\end{working}
What might be more surprising is that a similar program which discards the
pattern variables and instead uses the resources in the scrutinee is also
semantically linear, despite not being accepted by Linear Haskell:
\begin{notyet}
\begin{lstlisting}
f12 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* b *' $\lolli$ '* c) -> c
f12 x y use = case (x,y) of z { (a,b) -> use x y }
\end{lstlisting}
\end{notyet}
We justify that \incode{f12} is linear by appealing to the fact that the
expression (the tuple) being scrutinized is already in WHNF, evaluating it will
not consume neither $x$ nor $y$. Even if the tuple was constructed with
two expressions using $x$ and $y$ respectively, no computation would happen
since we aren't using neither $a$ nor $b$ (thereby never forcing the arguments
of the tuple). However, if we did use $a$ in the case body, then $x$ would be unavailable:
\begin{noway}
\begin{lstlisting}
f13 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* a *' $\lolli$ '* c) -> c
f13 x y use = case (x,y) of z { (a,b) -> use a x }
\end{lstlisting}
\end{noway}
This idea that $x$ and $a$ are mutually exclusive is the same behind let
bindings being mutually exclusive to the resources that define them.
By forcing the pattern variable (or the let binding), we run the computations
defined in terms of the linear variables used for that constructor argument (or
let binder body), but otherwise, if we don't use those binders, then we don't
run the computation thus no resources are consumed.

@@

Dually, consider the scrutinee to be an expression that's not in WHNF, s.t.
evaluating it to WHNF will require doing computation and thus consume linear
resources that are used in it:
\begin{notyet}
\begin{lstlisting}
f15 :: a *' $\lolli$ '* b *' $\lolli$ '* (a *' $\lolli$ '* b *' $\lolli$ '* (c,d)) -> (c,d)
f15 x y use = case use x y of z { (a,b) -> z }
\end{lstlisting}
\end{notyet}
Unlike when the scrutinee was in WHNF, we can no longer use $x,y$ in the case
alternatives, but we \emph{must} still use either the case binder $z$ or the linear
pattern variables $a,b$, e.g. it would be quite disastrous if any of the following typechecked:
\begin{noway}
\begin{lstlisting}
doubleFree :: Ptr *' $\lolli$ '* (Ptr *' $\lolli$ '* Result) -> Result
doubleFree x free = case free x of z { Result v -> free x }
\end{lstlisting}
\end{noway}
% Remover
\begin{noway}
\begin{lstlisting}
leakPointer :: Ptr -> ()
leakPointer x = case id x of z { _ -> () }
\end{lstlisting}
\end{noway}
%
%whereas this is fine:
%\begin{notyet}
%\begin{code}
%f16 :: Ptr -o (Ptr -o (Value,Int)) -> (Value,Int)
%f16 x free = case K (free x) of z { K y -> free x }
%\end{code}
%\end{notyet}
%
The result of evaluating the scrutinee must be consumed exactly to guarantee
that the resources used in the scrutinee are fully consumed, or risk them being only ``almost'' consumed. Take for example
\incode{use} in \incode{f15} to simply be \incode{(,)}: it is not sufficient for \incode{use x y} to be
evaluated to WHNF to consume \incode{x} and \incode{y}. Otherwise, if all the resources were
considered to be fully consumed after the scrutinee were evaluated in a case
expression, we could simply ignore the pattern variables, effectively
discarding linear resources (for cases such as the \incode{use = (,)} example). In
short, if the scrutinee is not in WHNF we must either consume the case binder
or the linear components of the pattern.

@@

%Remover estes
\begin{figure}[t]
  \begin{minipage}{0.5\textwidth}
\begin{noway}
\begin{lstlisting}
f w = case w of z
        (a,b) -> case (a,b) of z'
                        (c,d) -> (a,c)
\end{lstlisting}
\end{noway}
\vspace{-0.5cm}
\caption{Bad Case\label{fig:badcase}}
\end{minipage}%
\begin{minipage}{0.5\textwidth}
\begin{notyet}
\begin{lstlisting}
f w = case w of z
        (a,b) -> case (a,b) of z'
                        (c,d) -> (a,d)
\end{lstlisting}
\end{notyet}
\vspace{-0.5cm}
\caption{Good Case\label{fig:goodcase}}
\end{minipage}
\end{figure}

@@

%Trocar ordem com a metatheory
\section{Linear Core as a GHC Plugin\label{sec:discuss:implementation}}

@@

% The usage environments of a recursive group of binders (that is not necessarily
% strongly-connected) is computed in two separate passes.
% %
% First, we calculate a \emph{naive environment} by recording the linear
% resources semantically used in the binder bodies, while counting \emph{uses}
% (not syntactic occurrences) of the mutually recursive let variables.
% %
% Second, the binder names and corresponding \emph{naive environment} (mapping
% each linear resource to $1$ and the recursive variables to the $n$ number of
% times they are \emph{used}) in Cartesian pairs are given as input to
% % Algorithm~\ref{computeRecUsages},
% an algorithm which computes the actual usage environment
% of each binder.
%
% TODO: I should probably use the re-computed usageEnvs instead of the naiveUsageEnvs.
% I'm pretty sure it might fail in some inputs if I keep using the naiveUsageEnvs.
% \begin{algorithm}
% $usageEnvs \gets naiveUsageEnvs.map(fst)$\;
% \For{$(bind, U) \in naiveUsageEnvs$}{
%     \For{$V \in usageEnvs$}{
%         $V \gets sup(V[bind]*U\setminus\{bind\}, V\setminus\{bind\})$
%     }
% }
% \caption{computeRecUsages\label{computeRecUsages}}
% \end{algorithm}
%
% Intuitively, the algorithm, for each recursive binder, iterates over the
% (initially naive) usage environments and substitutes occurrences of the
% recursive binders by their corresponding usage environment,
% scaled up by the amount of times that recursive binder is used in the
% environment being updated. The result is the least upper bound of linear
% resources used by each strongly-connected group of mutually recursive bindings
% (for each binder of such a group), since all occurrences of the recursive
% binders in the usage environments will be substituted away by the corresponding
% recursive usage environments.
%
% The high-level description is:
% \begin{itemize}
%     \item Given a list of binders and their naive environment ($(f, F), (g, G),
%     (h, H)$) in which each use of $f, g, h$ is mapped to a count $n$ in $F, G, H$
%
%     \item For each pair of a letrec-bound variable and corresponding usage
%     environment, update all bindings and their usage as described in the
%     algorithm
%
%     \item After iterating through all bound rec vars, all usage environments
%         will be free of recursive bind usages, and hence describe the ``final'' usage environment
% \end{itemize}
%
% The complexity of computing a usage environment for $n$ recursive let binders
% is quadratic in $n$,
% % i.e. the algorithm has $O(n^2)$ complexity,
% but this is not an issue since it is uncommon to have more than a handful of
% binders in the same recursive let block.

@@

Programs ``linear modulo call-by-name'' are a class of programs
which scrutinize a variable, but then uses the variable in the case
alternatives of the case scrutinizing it. As will be made clear in
Section~\ref{sec:reverse-binder-swap-considered-harmful}, these programs can be
understood as linear as long as applications of linear functions binding these
variables are \emph{not} reduced \emph{call-by-name}, because, in doing so,
linear resources are duplicated. In Core, these programs are not seen as
linear, and thus GHC does not not reduce them using a call-by-name evaluation strategy.
% We have a flag in our implementation to accept some of these programs...
%
``Linear Rejected'' programs include different kinds of programs which are
rejected by the Linear Core system, but are still semantically linear. An
example is a program to which a rewrite rule was applied, resulting in an
application of unrestricted function to a linear resource.
% in the first argument
% to \incode{build}:
% \begin{lstlisting}
% take :: Int -> Replicator a *' $\lolli$ '* [a]
% take = \ (ds :: Int) (r :: Replicator b) ->
%      case ds of
%        1 -> build (\(c :: b -> a -> a) (n :: a) -> c (extract r) n)
%        ...
% \end{lstlisting}
% We know the linear resource is still used linearly because \incode{build} will
% always instance the unrestricted function to $(:)$ (read \emph{Cons}), which is
% linear. Additionally, these include programs scrutinizing a value of a type of
% which all constructors have no linear components, but only matching on the
% default alternative, programs where common-sub-expression elimination
% substituted more than one occurrence of \incode{Ur x} by \incode{y}, in the body of
% scrutinizing \incode{y} (a linear variable), and other programs affected by rewrite rules.
%
``Not Linear Rejected'' indicate programs that we do not understand as linear,
semantically, and are simultaneously rejected by Linear Core and its
implementation. These programs were not necessarily identified as bad
by the linter.\footnote{rever}
% An example, where the last component of \incode{HashMap}, \incode{wwz},
% is linear, but is not being consumed in the case alternative matching on it:
% \begin{lstlisting}
% jssvi :: Ur Bool *' $\lolli$ '* Set Int *' $\lolli$ '* Ur (TestT IO ())
% jssvi (a :: Ur Bool) (b :: Set Int)
%   = case a of
%      Ur ss -> case b of { HashMap wwx wwy wwz -> jump wjssAd ss }
% \end{lstlisting}
%
Finally, ``Unknown Rejected'' programs are those whose validity we did not
check. These include both programs accepted by Linear Core, but not by its
implementation, and programs that are simply rejected by Linear Core, but which
were not categorized.

@@

\section{Related and Future Work\label{sec:discussion}}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% {{{ Related Work
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%paragraph{Related Work}

% In this section we discuss related work, namely, Linear
% Haskell~\cite{cite:linearhaskell}, Linear Mini-Core~\cite{cite:minicore}, and
% linearity-influenced optimising
% transformations~\cite{cite:let-floating,peytonjones1997a,cite:linearhaskell}.

@@

\paragraph{Linear Haskell\label{sec:related-work-linear-haskell}.}
%
% \todo[inline]{Say something about Linear Haskell's lazy operational semantics,
% but the type system not being concerned with linearity in the presence of
% laziness}
%
% Haskell, contrary to most programming languages with linear types, has existed
% for 31 years of its life \emph{without} linear types. As such, the introduction
% of linear types to Haskell comes with added challenges that do not exist in
% languages that were designed with linear types from the start:
% %
% \begin{itemize}
%     \item Backwards compatibility. The addition of linear types shouldn't break
%         all existing Haskell code.
%     \item Code re-usability. The linearly-typed part of Haskell's ecosystem and
%         its non-linearly-typed counterpart should fit in together and it must be
%         possible to define functions readily usable by both sides
%         simultaneously.
%     \item Future-proofing. Haskell, despite being an
%         industrial-strength language, is also a petri-dish for experimentation
%         and innovation in the field of programming languages. Therefore, Linear
%         Haskell takes care to accommodate possible future features, in
%         particular, its design is forwards compatible with affine and dependent
%         types.
% \end{itemize}
% %
Linear Haskell~\cite{cite:linearhaskell} is concerned with retrofitting
linear types in Haskell but
is not concerned with extending GHC's intermediate language(s),
which presents its own challenges. 
%
Nonetheless, while the Linear Haskell work keeps Core unchanged, its
implementation in GHC does modify and extend Core with linearity/multiplicity
annotations. Core's type system is unable to type \emph{semantic} linearity of
programs (in contrast to \emph{syntactic} linearity), which results in Core
rejecting linear programs resulting from optimising transformations that
leverage the non-strict semantics of Core.
%
Linear Core overcomes the limitations of Core's linear type system derived from
Linear Haskell by understanding semantic linearity in the presence of laziness,
and provably accepts multiple Core-to-Core passes. Linear Core, ultimately, can
also be seen as a system that validates the programs written in Linear Haskell
and are compiled by GHC, by guaranteeing (through typing) that linear resources
are still used exactly once throughout the optimising transformations.
